Philosophy of DS\\ \large Best Practices for the Application of Computational Methods to Questions of Meaning

### 2017-06-01 pds/Paper/paper_pds.tex
final commit of day

The theory of DS has been used successfully in computational semantics - a branch of computational lingiustics. This recent success is attributable to increased availability of large amounts of digitalised language data and ML techniqeus.Particularly, it has become possible to obtain semantic representations immediately from natural language data. The theoretical framework of DS has become widely used and further developed as the advent and success of computational methods have enabled linguist to build large-scale models of linguistic meaning using digitalised corpora of text.

@@

DS is concerned with the specific topic of meaning in the study of natural language. If viewed from an application-oriented standpoint it could be seen as one of many sub-fields of  research. Indeed, as far as the present paper is concerned, it is precisely the application of DS in industry that warrants heightened attention.
From a philosophical standpoint, it is notoriously difficult to come to grips with the concept of meaning .
DS sits at the intersection of ethically touchy topics. 
First off, consider the situation where a meaning space is constructed based on some seemingly innocuous corpus and ends up containing severe bias along lines of gender or racial stereotypes. A recent study has found, for example, that a model trained on Google News articles display significant bias along binary gender dimensions \cite{bolukbasi2016man}. This raises a host of questions concerning the provenance of the bias, the adequate response by the responsible distributional semanticist and, not to mention, the nature of meaning.

Does one have to live with the fact that the meanings discovered in some corpus systematically contain unfavourable connotations for some group of people? After all, the corpus was produced by a particular society, so it should be expected that it mirror that society's prejudices. Is the stance "the model picked up what it is supposed to pick up, blame society" a fair one, or has the computational linguist the moral obligation to rid her model of biased meaning representations?

Fair and square, the model contains bias because the corpus contains bias. What if the semantic model is further used for some interaction within that same society that taught it - via corpus - what it means to be a \emph{woman}, will the bias be further re-enforced? It is has been reported that when one queries Google with the name of a person, then Google is more likely to show advertisement related to that person being arrested if the name's connotation is black \cite{sweeney2013discrimination}.
%A related issue is that even if results are valid, the of over-, or, respectively, under-representation of specific items in the data can disadvantage group.

\paragraph{The state of the ethical discussion in ML, NLP and DS}
For the domain general use of ML techniques there exists a series of studies by now from legal, political computer science experts which specify principles and guidelines. Furthermore, many non-technical articles have been published outside of academic channels. 
be specific, mention FATML. \\
\textbf{use nlp ethic article to get overview. do this shits.}

@@

\section{Guidelines for Machine Learning}\hypertarget{sec3}{}
\paragraph{Introduce FATML}
Algorithmically informed decisions are used in ever more domains of the private and public sectors and thus have the potential to impact society in significant ways. In many cases of undesirable effects that come with these technologies, legislation is behind in adequately protecting people negatively affected by it.

@@

The document then goes on to name five principles that should guide developders of algorithms in designing and implementing their systems. The authors of PAA urge creators to issue a Social Impact Statement wherein they detail the effect of their product has on society with respect to the five principles. In this way the principles are associated with a set of best practices.
\paragraph{5 Principles, quick recap of each of them}
We will now briefly go over the five principles - \emph{responsibility, explanability, accuracy, auditability and fairness} to describe their original intention in the PAA.
Responsibility requires developers to take seriously the possibility that adverse effects are produced their algorithmic system. Consequently, there should be designated roles within the organization to be responsible for redress if harm is caused. Excuses akin to ''the algorithm did it'' are no valid scapegoats. This reflects the acknowledgment that, at some level, there are people who design these systems and put them to use.
Explainability demands that both data selection and decisions taken by the algorithms can be explained to people affected by its usage in terms that are understandable by laywomen. The point here is not to obscure the workings of the systems to a point where no stakeholder can grasp it.
Accuracy asks designers and users of the system to keep track of its mistakes and sources of error, so that worst case scenarios become more easily anticipited and possible damage reduced.
Audibility is the requirement to create transparency in such a way that interested third parties are in the position to review the algorithms performance. This point is related to explainability.
Finally, Fairness requires that the algorithmic system does not affect different demographic groups in disparate and unfavourable ways, that is, discriminates against them.

\paragraph{What's the same}
It should be fair to say that some of these principles allow seamless transfer from general ML to any specific subdomain, as DS, without much further discussion.
This goes for the principles corresponding to accuracy - loggin mistakes is good practice no matter where -, auditibility - one should never obstruct uninterested parties' inquiries into the possibility of things going wrong - and explainability - obfuscating one's work and keeping affected parties ignorant is evidently bad practice both in ML and in DS.

\paragraph{What's different}
Much more interesting than the overlap lately mentioned is the question how the two remaining principles apply to to DS.
There are mainly two issues, one for each principle.

>>>>

The theory of DS has been used successfully in computational semantics - a branch of computational lingiustics. This recent success is attributable to increased availability of large amounts of digitalised language data and ML techniqeus. Particularly, it has become possible to obtain semantic representations immediately from natural language data. The theoretical framework of DS has become widely used and further developed as the advent and success of computational methods have enabled linguist to build large-scale models of linguistic meaning using digitalised corpora of text.

@@

DS is concerned with the specific topic of meaning in the study of natural language. If viewed from an application-oriented standpoint it could be seen as one of many sub-fields of NLP research. Indeed, as far as the present paper is concerned, it is precisely the application of DS in industry that warrants heightened attention.
For conreteness' sake, consider a situation where a meaning space is constructed based on some seemingly innocuous corpus and ends up containing severe bias along lines of gender or racial stereotypes. A recent study has found, for example, that a model trained on Google News articles displays significant bias along binary gender dimensions \cite{bolukbasi2016man}. This raises a host of questions concerning the provenance of the bias, the adequate response by the responsible distributional semanticist and, not to mention, the nature of meaning.

Does one have to live with the fact that the meanings discovered in some corpus systematically contain unfavourable connotations for some group of people? After all, the corpus was produced by a particular society, so it should be expected that it mirror that society's prejudices. Is the stance "the model picked up what it is supposed to pick up, blame society" a fair one, or has whoever works with the model the obligation to rid it of biased meaning representations? The questions go on: If the semantic model is further used for some interaction within that same society that taught it - via corpus - what it means to be a \emph{woman}, will the bias be further re-enforced? It is has been reported that when one queries Google with the name of a person, then Google is more likely to show advertisement related to that person being arrested if the name's connotation is black \cite{sweeney2013discrimination}.

\section{Ethics of Machine Learning}
\paragraph{The state of the ethical discussion in ML, NLP and DS}
For the domain general use of ML techniques there exists a series of studies by from legal, political computer science experts which specify principles and guidelines. Furthermore, many non-technical articles have been published outside of academic channels. 
be specific, mention FATML. \\
\textbf{use nlp ethic article to get overview. do this shits.}

@@

\section{Principles for  for Machine Learning}\hypertarget{sec3}{}
Algorithmically informed decisions are used in ever more domains of the private and public sectors and thus have the potential to impact society in significant ways. In many cases of undesirable effects that come with these technologies, legislation is behind in adequately protecting people negatively affected by it.

@@

The document then goes on to name five principles that should guide developders of algorithms in designing and implementing their systems. The authors of PAA urge creators to issue a Social Impact Statement wherein they detail the effect of their product has on society with respect to the five principles. In this way the principles are associated with a set of best practices.

We will now briefly go over the five principles - \emph{responsibility, explanability, accuracy, auditability and fairness} to describe their original intention in the PAA.
Responsibility requires developers to take seriously the possibility that adverse effects are produced their algorithmic system. Consequently, there should be designated roles within the organization to be responsible for redress if harm is caused. Excuses akin to ''the algorithm did it'' are no valid scapegoats. This reflects the acknowledgment that, at some level, there are people who design these systems and put them to use.
Explainability demands that both data selection and decisions taken by the algorithms can be explained to people affected by its usage in terms that are understandable by laywomen. The point here is not to obscure the workings of the systems to a point where no stakeholder can grasp it.
Accuracy asks designers and users of the system to keep track of its mistakes and sources of error, so that worst case scenarios become more easily anticipited and possible damage reduced.
Audibility is the requirement to create transparency in such a way that interested third parties are in the position to review the algorithms performance. This point is related to explainability.
Finally, Fairness requires that the algorithmic system does not affect different demographic groups in disparate and unfavourable ways, that is, discriminates against them.

It should be fair to say that some of these principles allow seamless transfer from general ML to any specific subdomain, as DS, without much further discussion.
This goes for the principles corresponding to accuracy - loggin mistakes is good practice no matter where -, auditibility - one should never obstruct uninterested parties' inquiries into the possibility of things going wrong - and explainability - obfuscating one's work and keeping affected parties ignorant is evidently bad practice both in ML and in DS.
Much more interesting than the overlap lately mentioned, however, is the question how the two remaining principles apply to to DS.
